Austral Spring Extratropical Southern Hemisphere zonally assymetric circulation \thanks{The research was supported by UBACyT20020170100428BA and the CLIMAX Project funded by Belmont Forum/ANR-15-JCL/-0002-01. Elio Campitelli was supported by a PhD grant from CONICET, Argentina.}

### 2022-05-31 analysis/paper/shceof.tex
REnderiza

\begin{abstract}
We analyse the joint variabilty of the zonally asymmetric springtime stratospheric and tropospheric circulation using Complex Empirical Orthogonal Functions (cEOF) to characterise planetary waves of varying amplitude and phase.
The leading cEOF represents variabilty of a zonal wave 1 in the stratosphere and a monopole in the troposphere that correlates slightly with the Symmetric Southern Annular Mode (S-SAM).
The second cEOF (cEOF2) is an alternative representation of the Pacific-South American modes.
One phase of this cEOF is also extremely correlated with the Asymmetric SAM (A-SAM) in the troposphere.
Springs with an active ENSO tend to phase-lock the cEOF2, but have no consistent impact on its magnitude.
Furthermore, we find indications that the location of Pacific Sea Surface Temperature anomalies affect the precise phase of the cEOF2.
\\
\keywords{
        Southern Hemisphere circulation \and
        Teleconnections \and
        Pacific South American Mode \and
        Southern Annular Mode \and
        Stratosphere \and
    }

@@

Another extensively used approach to characterize the SH tropospheric circulation anomalies, is by computing Empirical Orthogonal Functions (EOF, also known as Principal Component Analysis).
Within the EOF framework, the Southern Annular Mode (SAM) appears as the leading mode of variability of the SH circulation (Fogt and Marshall 2020).
SAM represents a relatively zonally symmetric pattern of alternating low pressures in polar latitude and a ring of high pressures in high latitudes with an embedded wave 3 pattern that is more prominent in the Pacific sector.
The 2nd and 3rd EOFs, usually known as Pacific--South American Patterns (PSA) 1 and PSA2 patterns, respectively, describe meridionally propagating wave trains that originate in the eastern equatorial Pacific and Australian-Indian Ocean sector, and travel towards the South Atlantic following a great-circle arch along the Antarctic coast (Mo and Paegle 2001).
These patterns influence precipitation anomalies in South America (Mo and Paegle 2001).
Although these patterns are usually derived by applying EOF to temporal anomalies, M. Raphael (2003) also applied EOF methods specifically to zonal anomalies.
More recently, Irving and Simmonds (2016) proposed a novel methodology for objectively identifying the PSA pattern using Fourier decomposition.

Patterns resulting from EOF analyses are more flexible than Fourier decomposition derived modes in the sense that they can capture oscillation patterns that cannot be characterised by purely sinusoidal waves with constant amplitude.
Nonetheless, they are restricted to standing oscillation modes and could not represent properly propagating or phase-varying features such as zonal waves.
A single EOF can also represent a mixture of two or more physical modes.

@@

An alternative methodology that has been proposed to study travelling and standing waves is complex Empirical Orthogonal Functions (Horel 1984).
This method extends EOF analysis to capture oscillations with varying amplitude and phase and has been applied to the time domain.
For instance, Krokhin and Luxemburg (2007) applied cEOF to station-based monthly precipitation anomalies and monthly temperature anomalies in the Eastern Siberia and the Far East region to characterise the main modes of variability and their connection to teleconnection indices.
Similarly, Gelbrecht, Boers, and Jürgen Kurths (2018) used cEOF applied to daily precipitation from reanalysis to study the propagating characteristics of the South American Monsoon.
To our knowledge, complex EOF has not been applied in the spatial domain to capture the phase-varying nature of planetary waves.

@@

We restrict this paper to the spring September-October-November (SON) trimester because this season experiences a maximum in tropical teleconnections over South America (Cazes-Boezio, Robertson, and Mechoso 2003) and Southern Hemisphere stratosphere-troposphere interactions (Lim, Hendon, and Thompson 2018).

In Section~\ref{methods} we describe the methods.
In Section~\ref{spatial} we analyse the spatial patterns of each complex EOF.
In Section~\ref{regressions} we study the spatial regressions with geopotential height, temperature and ozone anomalies.
In Section~\ref{psa} and \ref{sam} we analyse the relationship between cEOF2 the PSA and SAM modes.
In Section~\ref{tropical} we study tropical forcings that explain the variability of each cEOF.
In Section~\ref{precipitation} we show the relationship between these modes of variability and precipitation and surface temperature anomalies in South America and Oceania.
In Section~\ref{discussion} we compare our results with previous studies and discuss the benefits of our methodology.
In Section~\ref{conclusion} we summarise the conclusions.

@@

Preliminary analysis showed that the first cEOF was closely related to the zonal wave 1 of Total Ozone Column and the second cEOF was closely related to ENSO.
Therefore, to aid in the interpretation we chose the phase of cEOF1 so that the time series corresponding to the real part has the maximum correlation with the zonal wave 1 of Total Ozone Column between 75°S and 45°S.
Similarly, we chose the phase of cEOF2 so that the coefficient of determination between the Oceanic Niño Index (Bamston, Chelliah, and Goldenberg 1997) and the real part is minimized, which also nearly maximizes the correlation with the imaginary part.

@@

A significant upward trend in the real component of cEOF1 is evident (Fig.~\ref{fig:extended-series}a.1, p-value~= 0.0037) while there is no significant trend in any of the complex components of cEOF2.
The positive trend in the Real cEOF1 translates into a positive trend in cEOF1 magnitude, but not systematic change in phase (not shown).
This long-term change indicates an increase in the magnitude of the high latitude zonal wave 1.
In agreement, M. Raphael (2003) detected a step change after around 1975 in the temporal evolution of leading EOF computed from August-September-October 500 hPa zonal geopotential height anomalies, which is similar to the cEOF1.

@@

Figure~\ref{fig:eof1-regr-gh} shows regression maps of SON geopotential height anomalies upon cEOF1.
At 50 hPa (Figure~\ref{fig:eof1-regr-gh} row a), the Real cEOF1 is associated with a monopole centred over the Ross Sea.
The Imaginary cEOF1 is associated with a clear wave 1 pattern with maximum over the coast of East Antarctica.

At 200 hPa (Figure~\ref{fig:eof1-regr-gh} row b) the Real cEOF1 shows a single area of positive anomalies spanning West Antarctica surrounded by opposite anomalies in lower latitudes with is centre shifted slightly eastward compared with the upper level anomalies.
The Imaginary cEOF1 shows a much more zonally symmetrical pattern resembling the negative SAM phase (e.g. Fogt and Marshall 2020).
Both components of cEOF1 could be associated to zonal anomalies in the upper troposphere jet stream intensity.

Therefore, the magnitude and phase of the cEOF1 are associated with the magnitude and phase of a zonal wave only in the stratosphere.
While in the troposphere, they are associated with slightly off-centre monopoles.

@@

Figure~\ref{fig:eof2-regr-gh} shows the regression maps of geopotential height anomalies upon the cEOF2.
Wave trains similar to those identified for cEOF2 patterns (Fig~\ref{fig:ceofs-1}) can be distinguished.
Regressed anomalies associated with the Real cEOF2 are 90º out of phase with those associated with the Imaginary cEOF2.
All fields have a dominant zonal wave 3 limited to the western hemisphere, over the Pacific and Atlantic Oceans.
cEOF2 then represents an equivalent barotropic wave train that is very similar to the the PSA Patterns (Mo and Paegle 2001).
Comparing the location of the positive anomaly near 90ºW in column b of Figure~\ref{fig:eof2-regr-gh} with Figures 1.a and b from Mo and Paegle (2001), the Real cEOF2 regression map could be identified with PSA2, while the Imaginary cEOF2 resembles PSA1.

@@

The influence of cEOFs variability on air temperature was also evaluated.
Figure~\ref{fig:eof1-regr-t} shows regression maps of air temperature anomalies at 50 hPa and 200 hPa upon cEOF1.
In both levels, the Real cEOF1 is associated with a positive monopole over the South Pole with its centre moved slightly towards 150ºE (Fig.~\ref{fig:eof1-regr-t} column 1).
On the other hand, the regression maps on the Imaginary cEOF1 show a more clear wave 1 pattern with its maximum around 60ºE.
The distribution of temperature regression coefficients at 50 hPa and at 200 hPa mirror the geopotential height regression maps at 50 hPa (Fig.~\ref{fig:eof1-regr-gh}).

@@

Given the similarity between the cEOF2 regression maps (Fig.~\ref{fig:eof2-regr-gh}) and documented PSA patterns we study the relationship between these modes.
Table~\ref{tab:psa-eof2} shows the correlations between the two PSA indices and the time series for Real and Imaginary phases of cEOF2.
As anticipated by Figure~\ref{fig:eof2-regr-gh}, there is a strong positive correlation between PSA1 and Imaginary cEOF2, and also between PSA2 and Real cEOF2.
On the other hand, there is no significant relationship between PSA1 and Real cEOF2, and between PSA2 and Imaginary cEOF2.
As a result, cEOF2 represents well both the spatial structure and temporal evolution of the PSA modes, so it is possible to make an association between its two phases and the two PSA modes.
The same phase election for cEOF2 that maximises the relationship between ENSO and Imaginary cEOF2, also maximises the association between cEOF2 components and PSA modes (not shown).

@@

Figure~\ref{fig:phase-histogram} shows an histogram that counts the number of SON seasons in which the cEOF2 phase was close to each of the four particular phases (positive/negative of real/imaginary component), with the observations for each season marked as rugs on the horizontal axis.
In 62\% of seasons cEOF2 has a phase similar to either the negative or positive imaginary phase, making the imaginary phase the most common phase.
This is also, by construction, the phase that is most correlated with ENSO.
Therefore, the Imaginary cEOF2 explains more variance than the Real cEOF2 and conventional EOF analysis will tend to separate the two.

@@

We now explore the relationship between SAM and the cEOFs motivated by the resemblance between cEOFs regression maps and SAM patterns shown in Section~\ref{regressions}.
To that end, we computed the coefficient of determination between the cEOFs time series and the three SAM indices (SAM, A-SAM and S-SAM) defined by Campitelli, Díaz, and Vera (2022) at each vertical level (Fig.~\ref{fig:sam-eof-vertical}).
The SAM index is statistically significant correlated with the Real cEOF1 in all levels, and with the Imaginary cEOF1 and Imaginary cEOF2 in the troposphere.
On the other hand, correlations between SAM and the Real cEOF2 are non-significant.

@@

The connections between cEOFs and Tropical sources are also assessed, to explore plausible forcings for the described cEOFs patterns.
Figure~\ref{fig:sst-psi-2} shows the regression maps of Sea Surface Temperatures (SST) and streamfunction anomalies at 200 hPa respectively upon standardised cEOF2.
Besides the regression maps for the Real and Imaginary components, we include the corresponding regressions for two intermediate directions (corresponding to 45° and 135°).

Furthermore, the Imaginary cEOF2 (second row) is associated with strong positive SST anomalies on the Central Pacific and negative anomalies over an area across northern Australia, New Zealand the South Pacific Convergence Zone (SPCZ) (Figure~\ref{fig:sst-psi-2}.b1).
The regression field of SST anomalies bears a strong resemblance with canonically positive ENSO (Bamston, Chelliah, and Goldenberg 1997).
Indeed, there is a significant and very high correlation (0.76 (CI: 0.6 -- 0.87)) between the ONI and the Imaginary cEOF2 time series.
Besides the Pacific ENSO-like pattern, there are positive anomalies in the western Indian Ocean and negative values in the eastern Indian Ocean, resembling a positive IOD (Saji et al. 1999).
Consistently, the correlation between the Imaginary cEOF2 and the DMI is 0.62 (CI: 0.38 -- 0.78).

The Imaginary cEOF2 is associated with strong wave-like streamfunction anomalies emanating from the tropics (Figure~\ref{fig:sst-psi-2}.b2), both from the Central Pacific sector and the Indian Ocean.
These atmospheric response is consistent with the effect of ENSO and the IOD on the extratropics: the SST anomalies initiate anomalous tropical convection that excites Rossby waves that propagate meridionally towards higher latitudes (Mo 2000; Cai et al. 2011; Nuncio and Yuan 2015).

The cEOF2 is not associated with the same tropical SSTs for all their phases.
Figure~\ref{fig:sst-psi-2}.d1 and d2 show that the Real cEOF2 is not associated either with any significant SST nor streamfunction anomalies in the tropics.
As a result, the correlation between the Real cEOF2 and ENSO is also not significant (0 (CI: -0.3 -- 0.3)).
Meanwhile, Rows a and c in Fig.~\ref{fig:sst-psi-2} show that the intermediate phases are still associated with significant SST regressed anomalies over the Pacific Ocean, but with with slightly different location.
The 135º phase is associated with SST in the central Pacific (Fig.~\ref{fig:sst-psi-2}a.1), while the 45º phase is associated with SST in the eastern Pacific (Fig.~\ref{fig:sst-psi-2}c.1), which correspond roughly to the Central Pacific and Eastern Pacific ``flavours'' of ENSO, respectively (Kao and Yu 2009).
Both of these phases could be also associated to wave trains generated in the region surrounding Australia and propagates toward the extra-tropics.

@@

To better explore the relationship between tropical forcing and phases of the cEOF2, Figure~\ref{fig:enso-phase} shows the ONI index plotted against the cEOF2 phase for each SON trimester between 1979 and 2019, highlighting years in which the magnitude of cEOF2 is above the median.
In years with positive (negative) ONI, the cEOF2 phase is mostly around 90º (-90°), corresponding with positive (negative) Imaginary component.
In the neutral ENSO seasons, the cEOF2 phase is much more variable.
The black line in Figure~\ref{fig:enso-phase} is a sinusoidal fit of the relationship between ONI and cEOF2 phase.
The \(r^2\) corresponding to the fit is 0.57, statistically significant with p-value \textless{} 0.001, indicating a quasi-sinusoidal relation between these two variables.

The correlation between the absolute magnitude of the ONI index and the cEOF2 amplitude is 0.45 (CI: 0.17 -- 0.66).
However, this relationship is mostly driven by the three years with strongest ENSO events in the period (2015, 1997, and 1982) which coincide with the three years with strongest cEOF2 magnitude (not shown).
If those years are removed, the correlation becomes non-significant (0.04 (CI: -0.28 -- 0.35)).
Furthermore, even when using all years, the Spearman correlation --which is robust to outliers-- is also non-significant (0.2, p-value= 0.21).
Therefore, although the location of tropical SST seem to have an effect in defining the phase of the cEOF2, the relationship between the magnitude of cEOF2 and ONI remains uncertain.

@@

The influence of cEOFs variability in and 2-metre temperature and temperature in the Southern Hemisphere is also explored.
Figure~\ref{fig:pp-t2m-r2} shows the 2-meter temperature or precipitation anomalies explained variance by the multiple linear model of both Real and Imaginary cEOF1 components (column 1), and both Real and Imaginary cEOF2 components (column 2).
The variance explained by cEOF1 for precipitation anomalies is extremely low and also for most of the temperature anomalies, except for the eastern western of Antarctica (Fig.~ \ref{fig:pp-t2m-r2}a.1).

@@

Over South America, the Imaginary cEOF2 has positive correlations with precipitation in South Eastern South America (SESA) and central Chile, and negative correlations in eastern Brazil. This correlation field matches the springtime precipitation signature of ENSO (e.g. Cai et al. 2020) and it is also similar to the precipitation anomalies associated with the A-SAM (@ Campitelli, Díaz, and Vera 2022).

@@

\hypertarget{discussion}{%
\section{Discussion}\label{discussion}}

In previous sections, it has been shown that variability patterns that arise from cEOF methodology allow to describe the zonally asymmetric springtime extratropical SH circulation, reproducing previous features such as the variability related to PSAs or A-SAM.
However, it could be questioned what is the advantage of using the methodology proposed here.
Firstly, the spatial fields obtained from both components of cEOF2, that resembles PSA patterns, are in quadrature by construction.
Thus, it is possible to derive a joint PSA index from the amplitude and phase obtained when applying cEOF methodology.
Secondly, because of how these components are constructed, they are not forced to be orthogonal to the SAM as in the standard EOF methodology.
This allows us to show that Imaginary cEOF2, that resembles PSA1 variability, is closely related to the SAM in the troposphere.
Moreover, the high correlation of this component to the the zonally asymmetric portion of the SAM allow to speculate that the A-SAM might actually be a statistical contamination of the PSA mode.

Irving and Simmonds (2016) argued that there is some disagreement in the literature of whether the phase of the PSA pattern is affected by the location of tropical SST anomalies.
With the methodology used in this study, we were able to show not only that the cEOF2 tends to be in the positive or negative imaginary phase (\textasciitilde PSA1) when the ENSO region is warm or cold, respectively, but that central Pacific SST anomalies tend to align the cEOF towards the negative real phase and eastern Pacific SST anomalies tend to align it towards positive real phase.
When ENSO is neutral, the cEOF2 is still as active, but with no preferred phase.
This mirrors the results of Cai and Watterson (2002), who showed that the CSIRO Model can create PSA-like variability even in the absence of ENSO forcing (i.e.~with a climatological run), but that the variability of one of the PSA modes was enhanced when adding the ENSO signal.

The sensitivity of the phase of the PSAs to the location of the tropical SST anomalies was also seen by Ciasto, Simpkins, and England (2015), who detected similar Rossby wave patterns associated with central Pacific and eastern Pacific SST anomalies but with a change in phase.

Some results of our work contrast with those obtained by Irving and Simmonds (2016) when analysing the PSA.
They did not find neither a strong association between PSA-like variability and ENSO, nor a relationship between this variability and Australian precipitation.
Part of the disagreement could be due to the fact that we focus our analysis only on SON, which is the season when the teleconnection patterns are strongest, instead of the whole year.
Moreover, Irving and Simmonds (2016) use of ERA5 forectasted precipitation instead of the more observationally constrained CMAP dataset might be another source of disagreement.
Additionally, the discrepancy could be related to the fact that Irving and Simmonds (2016) show a weak relationship between ENSO and their PSA pattern, which explains partly the relationship between Australian rainfall and PSA-like variability.

\hypertarget{conclusion}{%
\section{Conclusion}\label{conclusion}}

In this study we assessed the Austral Spring Extratropical Southern Hemisphere zonally asymmetric circulation.
For this purpose, two complex indices were derived using Complex Empirical Orthogonal Functions.
This allowed us to characterise both the amplitude and phase of planetary waves that are not perfectly organised as sinusoidal waves with constant amplitude at each latitude circle.

>>>>

\begin{abstract}
We analyse the joint variabilty of the zonally asymmetric springtime stratospheric and tropospheric circulation using Complex Empirical Orthogonal Functions (cEOF) to characterise planetary waves of varying amplitude and phase.
The leading cEOF represents variabilty of a zonal wave 1 in the stratosphere and a monopole in the troposphere that correlates slightly with the Symmetric Southern Annular Mode (S-SAM).
The second cEOF (cEOF2) is an alternative representation of the Pacific-South American modes.
One phase of this cEOF is also extremely correlated with the Asymmetric SAM (A-SAM) in the troposphere.
Springs with an active ENSO tend to phase-lock the cEOF2, but have no consistent impact on its magnitude.
Furthermore, we find indications that the location of Pacific Sea Surface Temperature anomalies affect the phase of the cEOF2.
\\
\keywords{
        Southern Hemisphere circulation \and
        Teleconnections \and
        Pacific South American Mode \and
        Southern Annular Mode \and
        Stratosphere \and
    }

@@

Another extensively used approach to characterise the SH tropospheric circulation anomalies, is by computing Empirical Orthogonal Functions (EOF, also known as Principal Component Analysis).
Within the EOF framework, the Southern Annular Mode (SAM) appears as the leading mode of variability of the SH circulation (Fogt and Marshall 2020).
SAM represents a relatively zonally symmetric pattern of alternating low pressures in polar latitude and a ring of high pressures in high latitudes with an embedded wave 3 pattern that is more prominent in the Pacific sector.
The 2nd and 3rd EOFs, usually known as Pacific--South American Patterns (PSA) 1 and PSA2 patterns, respectively, describe meridionally propagating wave trains that originate in the eastern equatorial Pacific and Australian-Indian Ocean sector, and travel towards the South Atlantic following a great-circle arch along the Antarctic coast (Mo and Paegle 2001).
These patterns influence precipitation anomalies in South America (Mo and Paegle 2001).
Although these patterns are usually derived by applying EOF to temporal anomalies, M. Raphael (2003) also applied EOF methods specifically to zonal anomalies.
Irving and Simmonds (2016) proposed a novel methodology for objectively identifying the PSA pattern using Fourier decomposition.
More recently Goyal et al. (2022) created an index of amplitude and phase of zonal wave 3-like variability by combining the two leading EOFs of meridional wind anomalies.

Patterns resulting from EOF analyses are more flexible than Fourier decomposition derived modes in the sense that they can capture oscillation patterns that cannot be characterised by purely sinusoidal waves with constant amplitude.
Nonetheless, they are restricted to standing oscillation modes and could not properly represent propagating or phase-varying features such as zonal waves.
A single EOF can also represent a mixture of two or more physical modes.

@@

An alternative methodology that has been proposed to study travelling and standing waves is complex Empirical Orthogonal Functions {[}cEOF; Horel (1984){]}.
This method extends EOF analysis to capture oscillations with varying amplitude and phase and has been applied to the time domain.
For instance, Krokhin and Luxemburg (2007) applied cEOF to station-based monthly precipitation anomalies and monthly temperature anomalies in the Eastern Siberia and the Far East region to characterise the main modes of variability and their connection to teleconnection indices.
Similarly, Gelbrecht, Boers, and Jürgen Kurths (2018) used cEOF applied to daily precipitation from reanalysis to study the propagating characteristics of the South American Monsoon.
To our knowledge, cEOF analysis has not been applied in the spatial domain to capture the phase-varying nature of planetary waves in the atmosphere.

@@

We restrict this paper to the spring September-October-November (SON) trimester because this season experiences a maximum in tropical teleconnections over South America (Cazes-Boezio, Robertson, and Mechoso 2003) and SH stratosphere-troposphere interactions (Lim, Hendon, and Thompson 2018).

In Section~\ref{methods} we describe the methods.
In Section~\ref{spatial} we analyse the spatial patterns of each complex EOF.
In Section~\ref{regressions} we study the spatial regressions with geopotential height, temperature and ozone anomalies.
In Section~\ref{psa} and \ref{sam} we analyse the relationship between cEOF2 the PSA and SAM modes.
In Section~\ref{tropical} we study tropical forcings that explain the variability of each cEOF.
In Section~\ref{precipitation} we show the relationship between these modes of variability and precipitation and surface temperature anomalies in South America and Oceania.
In Section~\ref{discussion} we compare our results with previous studies and discuss the benefits of our methodology.

@@

Preliminary analysis showed that the first cEOF was closely related to the zonal wave 1 of Total Ozone Column and the second cEOF was closely related to ENSO.
Therefore, to aid in the interpretation we chose the phase of cEOF1 so that the time series corresponding to the real part has the maximum correlation with the zonal wave 1 of Total Ozone Column between 75°S and 45°S.
Similarly, we chose the phase of cEOF2 so that the coefficient of determination between the Oceanic Niño Index (Bamston, Chelliah, and Goldenberg 1997) and the real part is minimised, which also nearly maximises the correlation with the imaginary part.

@@

A significant positive trend in the real component of cEOF1 is evident (Fig.~\ref{fig:extended-series}a.1, p-value~= 0.0037) while there is no significant trend in any of the complex components of cEOF2.
The positive trend in the Real cEOF1 translates into a positive trend in cEOF1 magnitude, but not systematic change in phase (not shown).
This long-term change indicates an increase in the magnitude of the high latitude zonal wave 1.
In agreement, M. Raphael (2003) detected a step change after around 1975 in the temporal evolution of leading EOF computed from August-September-October 500 hPa zonal geopotential height anomalies, which is similar to the cEOF1.

@@

Figure~\ref{fig:eof1-regr-gh} shows regression maps of SON geopotential height anomalies upon cEOF1.
At 50 hPa (Figure~\ref{fig:eof1-regr-gh} row a), the Real cEOF1 is associated with a centre located over the Ross Sea.
The Imaginary cEOF1 is associated with a distinctive wave 1 pattern with maximum over the coast of East Antarctica.

At 200 hPa (Figure~\ref{fig:eof1-regr-gh} row b) the Real cEOF1 shows a single centre of positive anomalies spanning West Antarctica surrounded by opposite anomalies in lower latitudes, with is centre shifted slightly eastward compared with the upper-level anomalies.
The Imaginary cEOF1 shows a much more zonally symmetrical pattern resembling the negative SAM phase (e.g. Fogt and Marshall 2020).

Therefore, the magnitude and phase of the cEOF1 are associated with the magnitude and phase of a zonal wave mainly in the stratosphere.

@@

Figure~\ref{fig:eof2-regr-gh} shows the regression maps of geopotential height anomalies upon the cEOF2.
In the troposphere (Fig.~\ref{fig:eof2-regr-gh} column b), wave trains similar to those identified for cEOF2 patterns (Fig~\ref{fig:ceofs-1}) can be distinguished.
Regressed anomalies associated with the Real cEOF2 are 90º out of phase with those associated with the Imaginary cEOF2.
All fields have a dominant zonal wave 3 limited to the western hemisphere, over the Pacific and Atlantic Oceans.
cEOF2 then represents an equivalent barotropic wave train that is very similar to the the PSA Patterns (Mo and Paegle 2001).
Comparing the location of the positive anomaly near 90ºW in column b of Figure~\ref{fig:eof2-regr-gh} with Figures 1.a and b from Mo and Paegle (2001), the Real cEOF2 regression map could be identified with PSA2, while the Imaginary cEOF2 resembles PSA1.

These wave patterns are also present in the stratosphere (Fig.~\ref{fig:eof2-regr-gh} column a) supporting their equivalent barotropic nature.
But also present is a monopole over the pole with negative sign associated with the Real cEOF2 and positive sign associated with the Imaginary cEOF2.
This monopole might indicate strengthening of the polar vortex associated with positive values of the Real cEOF2 and weakening associated with positive Imaginary cEOF2.

@@

The signature of cEOFs variability on air temperature was also evaluated.
Figure~\ref{fig:eof1-regr-t} shows regression maps of air temperature anomalies at 50 hPa and 200 hPa upon cEOF1.
The distribution of temperature regression coefficients at 50 hPa and at 200 hPa mirror the geopotential height regression maps at 50 hPa (Fig.~\ref{fig:eof1-regr-gh}).
In both levels, the Real cEOF1 is associated with a positive centre over the South Pole with its centre moved slightly towards 150ºE (Fig.~\ref{fig:eof1-regr-t} column 1).
On the other hand, the regression maps on the Imaginary cEOF1 show a more clear wave 1 pattern with its maximum around 60ºE.

@@

Given the similarity between the cEOF2 related anomaly structures (Fig.~\ref{fig:eof2-regr-gh}) and documented PSA patterns we study the relationship between them.
Table~\ref{tab:psa-eof2} shows the correlations between the two PSA indices and the time series for Real and Imaginary phases of cEOF2.
As visually anticipated by Figure~\ref{fig:eof2-regr-gh}, there is a large positive correlation between PSA1 and Imaginary cEOF2, and between PSA2 and Real cEOF2.
On the other hand, there is no significant relationship between PSA1 and Real cEOF2, and between PSA2 and Imaginary cEOF2.
As a result, cEOF2 represents well both the spatial structure and temporal evolution of the PSA modes, so it is possible to make an association between its two phases and the two PSA modes.
That is, the phase election for cEOF2 that maximises the relationship between ENSO and Imaginary cEOF2, also maximises the association between cEOF2 components and PSA modes (not shown).

@@

Figure~\ref{fig:phase-histogram} shows an histogram that counts the number of SON seasons in which the cEOF2 phase was close to each of the four particular phases (positive/negative of real/imaginary component), with the observations for each season marked as rugs on the horizontal axis.
In 62\% of seasons cEOF2 has a phase similar to either the negative or positive imaginary phase, making the imaginary phase the most common phase.
This is also, by construction, the phase that is most correlated with ENSO.

Therefore, by virtue of being the most common phase, the Imaginary cEOF2 explains more variance than the Real cEOF2.
Conventional EOF analysis will therefore tend to separate them relatively cleanly, with the EOF representing the Imaginary cEOF2 always leading the one representing the Real cEOF2.

@@

We now explore the relationship between SAM and the cEOFs motivated by the resemblance between cEOFs regression maps and SAM patterns shown in Section~\ref{regressions}.
We then computed the coefficient of determination between the cEOFs time series and the three SAM indices (SAM, A-SAM and S-SAM) defined by Campitelli, Díaz, and Vera (2022) at each vertical level (Fig.~\ref{fig:sam-eof-vertical}).
The SAM index is statistically significant correlated with the Real cEOF1 in all levels, and with the Imaginary cEOF1 and Imaginary cEOF2 in the troposphere.
On the other hand, correlations between SAM and the Real cEOF2 are non-significant.

@@

The connections between cEOFs and Tropical sources of variability were also assessed.
Figure~\ref{fig:sst-psi-2} shows the regression maps of Sea Surface Temperatures (SST) and streamfunction anomalies at 200 hPa respectively upon standardised cEOF2.
Besides the regression maps for the Real and Imaginary components, we include the corresponding regressions for two intermediate directions (corresponding to 45° and 135°).

The Imaginary cEOF2 (second row) is associated with strong positive SST anomalies on the Central Pacific and negative anomalies over an area across northern Australia, New Zealand the South Pacific Convergence Zone (SPCZ) (Figure~\ref{fig:sst-psi-2}.b1).
The regression field of SST anomalies bears a strong resemblance with canonically positive ENSO (Bamston, Chelliah, and Goldenberg 1997).
Indeed, there is a significant and very high correlation (0.76 (CI: 0.6 -- 0.87)) between the ONI and the Imaginary cEOF2 time series.
Besides the Pacific ENSO-like pattern, there are positive anomalies in the western Indian Ocean and negative values in the eastern Indian Ocean, resembling a positive IOD (Saji et al. 1999).
Consistently, the correlation between the Imaginary cEOF2 and the DMI is 0.62 (CI: 0.38 -- 0.78).

The Imaginary cEOF2 is associated with strong wave-like streamfunction anomalies emanating from the tropics (Figure~\ref{fig:sst-psi-2}.b2), both from the Central Pacific sector and the Indian Ocean.
The atmospheric response associated with Imaginary cEOF2 is then consistent with the combined effect of ENSO and the IOD on the extratropics: with SST anomalies inducing anomalous tropical convection that in turn excite Rossby waves propagating meridionally towards higher latitudes (Mo 2000; Cai et al. 2011; Nuncio and Yuan 2015).

However, the cEOF2 is not associated with the same tropical SST anomaly patterns at all their phases
Figure~\ref{fig:sst-psi-2}.d1 and d2 show that the Real cEOF2 is not associated either with any significant SST nor streamfunction anomalies in the tropics.
As a result, the correlation between the Real cEOF2 and ENSO is also not significant (0 (CI: -0.3 -- 0.3)).
Meanwhile, Rows a and c in Fig.~\ref{fig:sst-psi-2} show that the intermediate phases are still associated with significant SST regressed anomalies over the Pacific Ocean, but at slightly different locations.
The 135º phase is associated with SST anomalies in the central Pacific (Fig.~\ref{fig:sst-psi-2}a.1), while the 45º phase is associated with SST anomalies (Fig.~\ref{fig:sst-psi-2}c.1) that correspond roughly to the Central Pacific and Eastern Pacific ``flavours'' of ENSO, respectively (Kao and Yu 2009).
Both phases could be also associated to wave trains generated in the region surrounding Australia and propagates toward the extra-tropics.

@@

To further explore the relationship between tropical forcing and phases of the cEOF2, Figure~\ref{fig:enso-phase} shows the ONI index plotted against the cEOF2 phase for each SON trimester between 1979 and 2019, highlighting years in which the magnitude of cEOF2 is above the median.
In years with positive (negative) ONI, the cEOF2 phase is mostly around 90º (-90°), corresponding with positive (negative) Imaginary component.
In the neutral ENSO seasons, the cEOF2 phase is much more variable.
The black line in Figure~\ref{fig:enso-phase} is a sinusoidal fit of the relationship between ONI and cEOF2 phase.
The \(r^2\) corresponding to the fit is 0.57, statistically significant with p-value \textless{} 0.001, indicating a quasi-sinusoidal relation between these two variables.

The correlation between the absolute magnitude of the ONI index and the cEOF2 amplitude is 0.45 (CI: 0.17 -- 0.66).
However, this relationship is mostly driven by the three years with strongest ENSO events in the period (2015, 1997, and 1982) which coincide with the three years with strongest cEOF2 magnitude (not shown).
If those years are removed, the correlation becomes non-significant (0.04 (CI: -0.28 -- 0.35)).
Furthermore, even when using all years, the Spearman correlation --which is robust to outliers-- is also non-significant (0.2, p-value= 0.21).
Therefore, although the location of tropical SST anomalies seem to have an effect in defining the phase of the cEOF2, the relationship between the magnitude of cEOF2 and ONI remains uncertain might be evident only in particularly intense ENSO events, for which there are few data.

@@

The influence of cEOFs variability in the anomalies of both 2-metre air temperature and precipitation in the Southern Hemisphere is also explored.
Figure~\ref{fig:pp-t2m-r2} shows the 2-meter temperature or precipitation anomalies explained variance by the multiple linear model of both Real and Imaginary cEOF1 components (column 1), and both Real and Imaginary cEOF2 components (column 2).
The variance explained by cEOF1 for precipitation anomalies is extremely low and also for most of the temperature anomalies, except for the northern tip of the Antarctic Peninsula, northern Weddell Sea and the Ross Sea coast (Fig.~ \ref{fig:pp-t2m-r2}a.1).

@@

Over South America, the Imaginary cEOF2 has positive correlations with precipitation in South Eastern South America (SESA) and central Chile, and negative correlations in eastern Brazil. This correlation field matches the springtime precipitation signature of ENSO (e.g. Cai et al. 2020) and it is also similar to the precipitation anomalies associated with the A-SAM (Campitelli, Díaz, and Vera 2022).

@@

\hypertarget{discussion}{%
\section{Discussion and conclusions}\label{discussion}}

In previous sections, it has been shown that variability patterns that arise from cEOF methodology describe the zonally asymmetric springtime extratropical SH circulation, reproducing previous features such as the variability related to PSAs or A-SAM.

However the advantage of using the methodology proposed here could be questioned.
Firstly, the spatial fields obtained from both components of cEOF2, that resembles PSA patterns, are in quadrature by construction.
Thus, the cEOF methodology allows to derive a joint PSA index from the resulted amplitude and phase.
Secondly, because of how these components are constructed, they are not forced to be orthogonal to other modes of circulation, as in the standard EOF methodology.
This allows us to show for example, that the Imaginary cEOF2, that resembles PSA1 variability, is closely associated to the SAM in the troposphere.
Moreover, the high correlation of this component to the zonally asymmetric portion of the SAM allow us to speculate that the A-SAM might actually be a statistical contamination of the PSA mode on the symmetric SAM.

Irving and Simmonds (2016) argued that there is some disagreement in the literature of whether the phase of the PSA pattern is affected by the location of tropical SST anomalies.
With the methodology used in this study, we were able to show not only that the cEOF2 tends to be in the positive or negative imaginary phase (\textasciitilde PSA1) when the ENSO region is warm or cold, respectively, but also that central Pacific SST anomalies tend to align the cEOF towards the negative real phase and eastern Pacific SST anomalies tend to align it towards positive real phase.
When ENSO phase is neutral, the cEOF2 is still as active, but with no preferred phase.
The latter agrees with the results of Cai and Watterson (2002), who showed that the CSIRO Model can develop PSA-like variability even in the absence of ENSO forcing (i.e.~with a climatological run), but that the variability of one of the PSA modes was enhanced when adding the ENSO signal.
The sensitivity of the phase of the PSAs to the location of the tropical SST anomalies was also seen by Ciasto, Simpkins, and England (2015), who detected similar Rossby wave patterns associated with central Pacific and eastern Pacific SST anomalies but with a change in phase.

Our method is similar to Goyal et al. (2022) as they also derive an index with amplitude and phase of wave-like activity in the SH with very similar patterns to our cEOF2.
Both indices are likely very similar (a detailed comparison is out of scope for this paper) but we believe that cEOF have the advantage of constructing the indices based on patterns that are exactly in quadrature by construction.
